---
title: "An Introduction to Lists"
author: "Colin Gillespie"
date: "`r Sys.Date()`"
vignette: >
   %\VignetteEngine{knitr::rmarkdown}
   %\VignetteIndexEntry{2. An Introduction to Lists}
output: knitr:::html_vignette
---

An R list is an object consisting of an ordered collection of (possibly
different) objects known as its components. For example, a list could consist of
a numeric vector and a data frame. To create a list, we use the `list` function

```{r}
lst = list(name = "Fred", wife = "Mary", children = 3, 
  child.ages = c(4, 7, 9))
``` 

Components are always numbered. In the `lst` object above, we can access
individual components using the double bracket notation. For example to select
the fourth element from the list, we have

```{r}
lst[[4]]
``` 

So to access the first element of the fourth component we
do

```{r}
lst[[4]][1]
``` 

The `length` function gives the number of top level components:

```{r}
length(lst)
``` 

Components of lists may also be named. If this is the case, then we
can also access the element using its name:

```{r}
lst[["wife"]]
lst$name
``` 

A useful function for interrogating an object is
`str`. Using the `lst` object
as an example, we have

```{r}
str(lst)
``` 
